Carrier phase and amplitude estimation for phase shift keying using pilots and data

### 2013-11-18 response.tex

\section*{Reviewer 1}

\begin{enumerate}
\item\textbf{Comment:} 
	This paper is concerned with carrier phase and amplitude estimation for PSK modulated signalling.
The considered least squares estimator seems to be only a minor extension of the one from [7]. Yet, more than 4 pages are dedicated to its derivation. Perhaps this section could be made more concise.
	\\\textbf{Response:}
We have shorteed Section~\zref{sec:least-squar-estim} but have left the description of the algorithm intact.  The primary reason for this is that much of the notation used during the analysis of the estimator is most naturally introduced during the description of the algorithm.  To qualify this for the reader we have included the following sentences at the beginning of Section~\zref{sec:least-squar-estim}.

``
In this section we derive Mackentun's algorithm to compute the least squares estimators of the carrier phase and amplitude~\cite{Mackenthun1994}.  Mackenthun specifically considered the non-data aided setting, so we modify the algorithm to include pilot symbols.  Although the modification is straightforward, much of the notation that we will use for analysing the estimator in Section~\ref{sec:stat-prop-least} is naturally introduced during the description of the algorithm.  For this reason, and so that our paper is self contained, we describe the algorithm in full.
''

%	We have shortened Sec.\ II to make it more concise and improve readability.
\item\label{R1it2}\textbf{Comment:} 
	As stated by the authors themselves, the key results and main contributions of the paper involve the asymptotic properties of the least squares estimator. As far as I can see the derivation seems correct and complete, yet it is mathematically very involved and rather difficult to follow.
	\\\textbf{Response:}	
	For improved readability of the manuscript, we have now moved all of the lemmas into the appendix. We have also added a discussion of the implications and the practical relevance of the theorems after the statement of Theorems~\zref{thm:consistency} and~\zref{thm:normality} in Section.~\zref{sec:stat-prop-least}.  We have also expanded the introduction of the manuscript to make the practical and theoretical importance of our of results clear.

@@

\item\textbf{Comment:} 
	By detailed inspection of Fig.\ 4 showing the jitter variance for amplitude estimates, it is easily observed that $|P| < L$ performs better than $|P| = L$ in the medium-to-low SNR range. In my mind this is somewhat strange, since it should be the other way round (similar to the evolution of the phase curves in Figs.\ 1Ð3 and 5). Can you comment on this phenomenon ?
	\\\textbf{Response:}
In the original manuscript we plotted the \emph{variance} of the amplitude estimator, not the mean square error. That is, the solid lines where the theoretical variance of $\hat{m}_L = \hat{\rho} - \rho_0 G(0)$ and each point from the Monte-Carlo simulations was computed as the unbiased error $\tfrac{1}{T}\sum_{i=1}^T\big(\hat{\rho}_i - \rho_0G(0)\big)^2$.  We chose this approach in order to highlight the correctness of our asymptotic theory. This was decribed in the 4th paragraph of the old Section~8 containing the simulations section.

We do agree with the reviewer this this plot by itself initially appears misleading.  To aleviate in the new manuscript we instead plot the average square error computed in the usual way by Monte-Carlo simulation and the solid line is given by the bias plus variance squared.  The following paragraph has also be placed in the simulations section (Section~\zref{sec:simulations}) in the new manuscript.

``Figure~\zref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
shows the mean square error of the amplitude estimator when $M=2$ and $L=2048$ and when the number of pilots symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the bias $\rho_0\big(1 - G(0)\big)$ squared plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\zref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\]
When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases, so the bias is significant when SNR is small.  Figure~\zref{fig:plotampBPSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.
''

\item\textbf{Comment:} 
	In a much more powerful approach to the topic addressed in the current paper, the pilot sequence would be employed for a coarse estimation/correction of the carrier phase of the payload data. In the sequel, by embedding carrier phase recovery into an iterative (turbo) decoding scheme [d, e] or any other error correcting method [f, g], the jitter performance is close to the CRLB for DA estimation even at very small SNRs, which are mainly determined by the convergence region of the (turbo) decoder. By taking into account the fact that modern and powerful communication systems are not conceivable without any FEC scheme, it is not clear to me why I should choose your algorithm in such a context.
	\\\textbf{Response:}
	We have added the following paragraph to the introduction describing the need for sufficiently accurate coarse estimates to initialise an iterative code-aided synchronisation stage. %citing [d, e].

@@

\section*{Reviewer 3}

\begin{enumerate}
\item
	\textbf{Comment:} 
The paper studies the estimation of amplitude and phase due to the distortion introduced by the channel in communication systems with $M$-PSK signals. The estimator is derived from the observation of $L$ noisy samples taken at a rate of one sample per symbol. Part of the $L$ observed samples corresponds to pilot transmitted symbols, that are already known a priori by the receiver, while the rest of samples are associated with data symbols which are unknown at the receiver. The focus of the article is to determine the amplitude and phase without considering the estimation of the transmitted data like this problem is usually treated in the literature. The work modifies the Mackenthun algorithm to include samples corresponding to pilot symbols while the Mackenthun paper calculates a least squares estimator for a similar problem but without pilot symbols.

The major contribution of this paper is the description of the asymptotic properties of the least squares estimator for the phase and amplitude. Although it seems technically correct, the authors do not make clear the relevance of their work to signal processing applications and to timely problems. They limit themselves in presenting several lemmas and proofs. Therefore, I cannot recommend this manuscript for publication in the IEEE Transactions on Signal Processing.
	\\\textbf{Response:}
We thank their reviewer for the comments. We have greatly expanded the introduction of the manuscript to make the practical and theoretical importance of our of results clear.
\end{enumerate}

### 2013-11-18 paper.tex
modified the mse amplitude estimator plot

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimators of carrier phase and amplitude asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under comparatively mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of a non data-aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convinving response is that, while the specific results given here apply only to the least square estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection} in the literature~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.

@@

That this paper contains novel theoretical developments should does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares carrier phase and amplitude estimators can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing systems parameters such as the number and placement of pilot symbols and also provides useful information to other reciever components, such as channel tracking systems.

@@

Figure~\ref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
shows the variance of the amplitude estimator when $M=2$ and with $L=32, 256, 2048$ and when the number of pilots symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines are the variance predicted by Theorem~\ref{thm:normality}.  The dots and crosses show the results of Monte-Carlo simulations.  Each point is computed as the unbiased error $\tfrac{1}{T}\sum_{i=1}^T\big(\hat{\rho}_i - \rho_0G(0)\big)^2$.  This requires $G(0)$ to be known.  In practice $G(0)$ may not be known at the receiver, so Figure~\ref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
serves to validate the correctness of our asymptotic theory, rather than to suggest the practical performance of the amplitude estimator.  When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases, so the bias is significant when SNR is small.  %As indicated in Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} the variance of the least squares amplitude estimator is smaller than that of the estimator when all symbols are pilots.  However, due to the bias, the MSE of the least squares amplitude estimator is \emph{not} smaller than that of the unmodulated carrier.

Figure~\ref{fig:plotphaseQPSKmultL} shows the MSE of the phase estimator when $M=4$ and $L=32,256, 2048$ and the number of pilots is $\abs{P}=\tfrac{L}{8},L$.  The figure depicts an interesting phenomenon.  When $L=2048$ and $\abs{P} = \tfrac{L}{8} = 256$ the number of pilots symbols is the same as when $L=\abs{P} = 256$.  When the SNR is small (approximately less than \unit[0]{dB}) the least squares estimator using the $256$ pilots symbols and also the $2048-256=1792$ data symbols performs \emph{worse} than the estimator that uses only the $256$ pilots symbols.  A similar phenomenon occurs when $L=256$ and $\abs{P} = \tfrac{L}{8} = 32$.  %When the SNR is small it is better to ignore the data symbols and use only the pilot symbols.  
% This behaviour suggests modifying the objective function to give the pilots symbols more importance when the SNR is low.  For example, rather than minimise~\eqref{eq:SSdefn} we could instead minimise a weighted version of it,
% \[
% SS_{\beta}(a, \{d_i, i \in D\}) = \sum_{i \in P} \abs{ y_i - a s_i }^2 + \beta \sum_{i \in D} \abs{ y_i - a d_i }^2,
% \]
% where the weight $\beta$ would be small when SNR is small and near $1$ when SNR is large.  Computing the $\hat{a}$ that minimises $SS_\beta$ can be achieved with only a minor modification to algorithm~\ref{alg:loglinear}.  Line~\ref{alg_gset} is modified to $g_i = \beta y_i e^{-j u}$ and lines~\ref{alg_hata1}~and~\ref{alg_hata2} are modified to $\hat{a} =  \frac{1}{\abs{P} + \beta\abs{D}} Y$.  For fixed $\beta$ the asymptotic properties of this weighted estimator could be derived using the techniques we have developed in Sections~\ref{sec:stat-prop-least},~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm}.  This would enable a rigorous theory for selection of $\beta$ at the receiver.  One caveat is that the receiver would require knowledge about the noise distribution in order to advantageously choose $\beta$.  We do not investigate this further here.

@@

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM2-1.mps}
		\caption{Amplitude estimator variance (bias removed) versus SNR for BPSK.}
		\label{fig:plotampBPSK}
\end{figure}

>>>>

In this paper we present an alternative approach to the study of estimators for carrier phase and amplitude.  Rather than compute bounds, we directly study the least squares estimator asymptotically as the number of transmitted symbols $L\to \infty$.  We show, under comparatively mild conditions on the distribution of the noise, that the least squares estimator of carrier phase is strongly consistent (asymptotically unbiased) and asymptotically normally distributed.  However, the least squares amplitude estimator is asymptotically biased and converges to a positive real number that is a function of the true carrier amplitude, the noise distribution and the size of the constellation $M$.  To our knowledge, this is the first time an analytic description of the bias or variance of a non data-aided amplitude estimator has been obtained.  Our expressions for the asymptotic variance and bias are simple to compute.  %Our results are indirectly supported by the simulation study of Pauluzzi~\cite{Pauluzzi2000}.  We present our our set of simulations supporting our analytic results in Section~\ref{}.

@@

One criticism that might be directed at our asymptotic approach when compared to the Cramer-Rao bounds already developed in the literature is that our specific results apply only to the least square estimator.  One response to this criticism is that the least squares estimator should be a popular practical choice because it can be efficiently computed in $O(L\log L)$ operations and because it is the maximum likelihood estimator under the common assumption that the noise is white and Gaussian.  Perhaps an even more convinving response is that, while the specific results given here apply only to the least square estimator, the techniques that we develop to derive these results are now able to be deployed for the analysis of other estimators.  For example, in practical systems it is sometimes desirable to place more importance on certain data symbols and use a small constellation size, say BPSK ($M=2$) for these, but to use a larger constellation size, say QPSK ($M=4$), for other less important symbols.  This is related to what is called \emph{unequal error protection}~\cite{Aydinlik_turbo_uep_2008,Sandberg_uep_ldpc_2010}.  It would be reasonably straigthforward to extend our results to allow data and pilot symbols with varying constellation size.  A more involved extension would be the inclusion of carrier frequency as a parameter to be estimated.  The techniques we present, combined with those from~\cite{McKilliamFrequencyEstimationByPhaseUnwrapping2009,McKilliam_mean_dir_est_sq_arc_length2010}, would potentially allow this.

@@

That this paper contains novel theoretical developments does not imply that our results are of no immediate practical interest.  The design of our estimator was motivated by a project at the University of South Australia's Institute for Telecommunications Research to develop a narrowband multiuser communication system to collect sensor network data using a satellite in orbit.  This estimator has been implemented on a prototype that was successfully trialled from April to July of 2013 with signals from multiple low power ground-based transmitters being successfully received and decoded by both aircraft and satellites in orbit~\cite{ASRPpromovideo}.  Our theoretical analysis aided design of this system.  Based on our Theorem~\ref{thm:normality} in Section~\ref{sec:stat-prop-least} the performance of the least squares estimator can be accurately predicted without the need for time consuming Monte-Carlo simulations.  This provides easy to use and rigourously justified methods for choosing systems parameters, such as the number and placement of pilot symbols, and also provides useful information to other reciever components, such as channel tracking systems.

@@

% Figure~\ref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
% shows the variance of the amplitude estimator when $M=2$ and with $L=32, 256, 2048$ and when the number of pilots symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines are the variance predicted by Theorem~\ref{thm:normality}.  The dots and crosses show the results of Monte-Carlo simulations.  Each point is computed as the unbiased error $\tfrac{1}{T}\sum_{i=1}^T\big(\hat{\rho}_i - \rho_0G(0)\big)^2$.  This requires $G(0)$ to be known.  In practice $G(0)$ may not be known at the receiver, so Figure~\ref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
% serves to validate the correctness of our asymptotic theory, rather than to suggest the practical performance of the amplitude estimator.  When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases, so the bias is significant when SNR is small.  %As indicated in Figures~\ref{fig:plotampBPSK},~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} the variance of the least squares amplitude estimator is smaller than that of the estimator when all symbols are pilots.  However, due to the bias, the MSE of the least squares amplitude estimator is \emph{not} smaller than that of the unmodulated carrier.

Figure~\ref{fig:plotampBPSK} %,~\ref{fig:plotampQPSK}~and~\ref{fig:plotamp8PSK} 
shows the mean square error of the amplitude estimator when $M=2$ and $L=2048$ and when the number of pilots symbols is $\abs{P} = 0, \tfrac{L}{2}, L$.  The solid lines show the mean square error predicted by the bias $\rho_0\big(1 - G(0)\big)$ squared plus the variance of $\hat{m}_L = \hat{\rho} - \rho_0G(0)$ predicted by Theorem~\ref{thm:normality}.  That is, the solid lines are given by
\[
\rho_0^2\big(1 - G(0)\big)^2 + \rho_0^2\frac{p B_1 + d B_2}{L}. 
\]
When SNR is large $G(0)$ is close to $1$ and the bias of the amplitude estimator is small.  However, $G(0)$ grows without bound as the variance of the noise increases, so the bias is significant when SNR is small.  Figure~\ref{fig:plotampBPSK} suggests our asymptotic theory to be an excellent predictor of the mean square error of the amplitude estimator.

Figure~\ref{fig:plotphaseQPSKmultL} shows the MSE of the phase estimator when $M=4$ and $L=32,256, 2048$ and the number of pilots is $\abs{P}=\tfrac{L}{8},L$.  The figure depicts an interesting phenomenon.  When $L=2048$ and $\abs{P} = \tfrac{L}{8} = 256$ the number of pilots symbols is the same as when $L=\abs{P} = 256$.  When the SNR is small (approximately less than \unit[0]{dB}) the least squares estimator using the $256$ pilots symbols and also the $2048-256=1792$ data symbols performs \emph{worse} than the estimator that uses only the $256$ pilots symbols.  A similar phenomenon occurs when $L=256$ and $\abs{P} = \tfrac{L}{8} = 32$.  %When the SNR is small it is better to ignore the data symbols and use only the pilot symbols.   
% This behaviour suggests modifying the objective function to give the pilots symbols more importance when the SNR is low.  For example, rather than minimise~\eqref{eq:SSdefn} we could instead minimise a weighted version of it,
% \[
% SS_{\beta}(a, \{d_i, i \in D\}) = \sum_{i \in P} \abs{ y_i - a s_i }^2 + \beta \sum_{i \in D} \abs{ y_i - a d_i }^2,
% \]  
% where the weight $\beta$ would be small when SNR is small and near $1$ when SNR is large.  Computing the $\hat{a}$ that minimises $SS_\beta$ can be achieved with only a minor modification to algorithm~\ref{alg:loglinear}.  Line~\ref{alg_gset} is modified to $g_i = \beta y_i e^{-j u}$ and lines~\ref{alg_hata1}~and~\ref{alg_hata2} are modified to $\hat{a} =  \frac{1}{\abs{P} + \beta\abs{D}} Y$.  For fixed $\beta$ the asymptotic properties of this weighted estimator could be derived using the techniques we have developed in Sections~\ref{sec:stat-prop-least},~\ref{sec:proof-almost-sure} and~\ref{sec:proof-asympt-norm}.  This would enable a rigorous theory for selection of $\beta$ at the receiver.  One caveat is that the receiver would require knowledge about the noise distribution in order to advantageously choose $\beta$.  We do not investigate this further here.

@@

\begin{figure}[tp]
	\centering
		\includegraphics[width=\linewidth]{code/data/plotM2-1.mps}
		\caption{Amplitude estimator versus SNR for BPSK with $L=2048$}
		\label{fig:plotampBPSK}
\end{figure}
